### 2010-07-22 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: 'replace' fits in single lines
Desc@ICFP: alternative to switchD: using fsigma

In order to capture a notion of higher-order induction, we add a code
$\SYMBDHindx$ that takes an indexing set $\V{H}$. This amounts to give
a recursive subobject for each element of $H$.
%
\[\stk{
    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Set \smallskip \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     \mapsto \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
}\]

@@

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be.
We eliminate the recursion by \emph{specialising} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
                \EnumT{\V{E}} \To \Desc
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together the declaration
of \(\Desc\) and its internal copy, we have made it appear in its own
type. Hardwired as a \emph{fait accompli}, this creates no regress,
although one must assume the definition to recheck it.\footnote{Our
  Agda model does not formalize the $\SYMBswitchD$
  construction. However, it proves the isomorphism between the defined
  descriptions and the coded ones. The role of $\SYMBswitchD$ is to
  collapse this isomorphism, hence identifiying defined and coded
  descriptions.}

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a readily propagated non-dependent return type $\V{T}$.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
% 
We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). $\V{xs}$ is of the right shape, but its
sub-elements are of the wrong type. On the other hand, for each
sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding element
in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
program to do that---please forgive us if we lapse to a pattern
matching notation, for readability's sake.
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}             \mapsto
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} \mapsto        \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
}\]
%
Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.

>>>>

In order to capture a notion of higher-order induction, we add a code
$\SYMBDHindx$ that takes an indexing set $\V{H}$. This amounts to give
a recursive subobject for each element of $H$.
%
\[\stk{
    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Desc
    \smallskip \\
    \descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     
        \mapsto 
            \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
}\]

@@

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be.
We eliminate the recursion by \emph{specialising} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
                \EnumT{\V{E}} \To \Desc
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together the declaration
of \(\Desc\) and its internal copy, we have made it appear in its own
type. Hardwired as a \emph{fait accompli}, this creates no regress,
although one must assume the definition to recheck it\footnote{Our
  Agda model does not formalize the $\SYMBswitchD$
  construction. However, it proves the isomorphism between the defined
  descriptions and the coded ones. The role of $\SYMBswitchD$ is to
  collapse this isomorphism, hence identifiying defined and coded
  descriptions.}. Alternatively, we could have extended $\Desc$ with a
code for \emph{finite} sums, internalizing the $\SYMBswitch$ in the
interpretation function:
%%
\[\stk{
    \Dsigma{(\Bhab{\V{E}}{\EnumU})}
           {(\Bhab{\V{B}}{\spi{\V{E}}{\LAM{\_}{\Desc}}})} : \Desc
    \smallskip \\
    \descop{\Dsigma{\V{E}}{\V{B}}}{\V{X}}     
        \mapsto 
            \SIGMAS{\Bhab{\V{x}}{\EnumT{\V{E}}}}
                   {\descop{\switch{\V{E}}{(\LAM{\_}{\Desc})}{\V{B}}{\V{x}}}{\V{X}}}    
}\]
%%
Using $\SYMBDsigma$ to switch over the $\Desc$ constructors, we avoid
mention of $\Desc$ while levitating it. The $\SYMBswitch$ operator is
pushed into the interpretation function, at which point $\Desc$ indeed
exists.

%% * There exists alternative
%% **   <- All we want is to avoid mention of Desc
%% **   -> Add a code for finite sigma : $\sigma : (e : \EnumU) -> \pi e \Desc -> \Desc$
%% **       <- Take a finite set in argument
%% **       <- Interpreted as a switch
%% **       -> No more need to mention Desc during its construction
%% **           <- Pushed into intrepretation, at which time it exists

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a readily propagated non-dependent return type $\V{T}$.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
% 
We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). $\V{xs}$ is of the right shape, but its
sub-elements are of the wrong type. On the other hand, for each
sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding element
in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
program to do that---please forgive us if we lapse to a pattern
matching notation, for readability's sake.
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void   &
    \Void  \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}  &
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} &
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} &
 \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
\end{array}
}\]
%
Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.
